A linear inside-outside algorithm for correcting sequencing errors in structured RNA sequences

### 2012-10-07 Recomb/methods_RECOMB.tex
Complexity chapter ... at last!

\subsubsection{Combining Inside and Outside value into base content probabilities}
By construction, the partition function over all sequences at exactly $m$ mutations of $s$ can 
be described in function of the \emph{forward} term as $\Z{0,n-1}{m}{X,X}$,
 for any nucleotide $X\in B$ or
in function of the \emph{backward} term, for any unpaired position $k$:
$$
	\Z{0,n-1}{m}{X,X}
	\equiv
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a}
$$

We are now left to compute the probability that a given position is a given nucleotide.
We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of allowed mutations. 
\begin{align*}
	\mathbb{P}(s_i = x\mid s,\Omega, S,M) &:= \frac{\mathcal{B}(i,x,s,\Omega,S,M)}{\sum_{m=0}^{M}\Z{0,n-1}{m}{X,X}}\\ 
\mathcal{B}(i,x,s,\Omega,S,M)&=
 \left\{
	\begin{array}{ll}
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		&\text{If }S_i = -1\\
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{-E_{(i,k),\varnothing\to xb}^{\Omega,\beta} }{RT}}
			\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\Z{i+1,k-1}{m'}{x,b}
		&\text{If }S_i=k>i\\
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{\frac{-E^{\Omega, \beta}_{(k,i),\varnothing\to bx}}{RT}}
			\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\Z{k+1,i-1}{m'}{b,x}
		&\text{If }S_i=k<i
	\end{array}\right.
\end{align*}

@@

>>>>

\subsubsection{Combining Inside and Outside values into point-wise mutations probabilities}
By construction, the partition function over all sequences at exactly $m$ mutations of $s$ can 
be described in function of the \emph{forward} term as $\Z{0,n-1}{m}{X,X}$,
 for any nucleotide $X\in B$ or
in function of the \emph{backward} term, for any unpaired position $k$:
$$
	\Z{0,n-1}{m}{X,X}
	\equiv
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a}
$$

We are now left to compute the probability that a given position is a given nucleotide.
We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of allowed mutations. 
\begin{align}
	\mathbb{P}(s_i = x\mid s,\Omega, S,M) &:= \frac{\mathcal{W}(i,x,s,\Omega,S,M)}{\sum_{m=0}^{M}\Z{0,n-1}{m}{X,X}}\label{eq:normalize}\\ 
\mathcal{W}(i,x,s,\Omega,S,M)&=
 \left\{
	\begin{array}{ll}
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		&\text{If }S_i = -1\\
			\displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{-E_{(i,k),\varnothing\to xb}^{\Omega,\beta} }{RT}}
			\cdot\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\cdot\Z{i+1,k-1}{m'}{x,b}
		&\text{If }S_i=k>i\\
    \displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{\frac{-E^{\Omega, \beta}_{(k,i),\varnothing\to bx}}{RT}}
			\cdot\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\cdot\Z{k+1,i-1}{m'}{b,x}
		&\text{If }S_i=k<i
	\end{array}\right.\label{eq:combine}
\end{align}

@@

\subsection{Complexity}
Equations~\eqref{eq:Z_rec} and~\eqref{eq:Y_rec} can be computed using dynamic programming, i.e. $\mathcal{Z}^{*}_{*}$ or $\mathcal{Y}^{*}_{*}$ terms are computed starting from smaller values of $m$ and interval lengths, memorizing the results such that they become available in constant time during future stages of the computation. Furthermore, energy terms $E(\cdot)$ can be accessed in constant time (after a precomputation in $\Theta(n\cdot|\Omega|)$ for the isostericity part). Computing any given term therefore requires $\Theta(m)$ operations.

In principle, this action should be performed $\Theta(m\cdot n^2)$ times to investigate any triplet $(m,i,j)$.
However, a close inspection of the recurrences reveal that the computation can be safely restricted to a subset of all intervals.
For instance, in the forward computation, such intervals $[i,j]$ do not break any base-pair, and their next position $j+1$ is either past the end of the sequence, or is base-paired prior to $i$. Similar constraints hold for the outside computation. Such constraints drastically limit the combinatorics of required computations, dropping from $\Theta(n^2)$ to $\Theta(n)$ the number of terms that need to be computed and stored. Consequently the overall complexity of the algorithm is $\Theta(n\cdot(|\Omega|+m^2))$ aritmetic operations and $\Theta(n\cdot(|\Omega|+m))$ memory.

### 2012-10-07 Recomb/results_RECOMB.tex
Complexity chapter ... at last!

The time benchmarks were done on a MacMini 2010, 2.3GHz dual-core Intel Core i5, 8GB of RAM.
Since applications of \RNApyro implies a need for
 efficiency and scalability. In Table~\ref{tab:time} we
present different times needed to compute the probabilities for  every nucleotide at every positions for a vast set of parameters. For those test
 all the sequences were generated randomly, as the target secondary structure.

@@

We build our data set from the 5S rRNA multiple sequence alignment (MSA) available in the Rfam Database 11.0 (Rfam id: \texttt{RF00001}).
Since our software does not currently implement gaps (mainly because scoring indels is a challenging issue that cannot be fully addressed
in this work),  we clustered together the sequences with identical gap locations. From the $54$ MSAs without gap produced, we selected the
biggest MSA  which contains $130$ sequences (out of $712$ in the original Rfam MSA). Then, in order to avoid any bias, we used \texttt{cd-hit}
\cite{Li:2006fk} to remove sequences with more than 80\% of sequence similarity. This operation resulted in a data set of $45$ sequences.

>>>>

The time benchmarks were done on a MacBook Pro 2010, 2 processors at 2.66GHz, 8GB of RAM.
Since applications of \RNApyro implies a need for
 efficiency and scalability. In Table~\ref{tab:time} we
present different times needed to compute the probabilities for  every nucleotide at every positions for a vast set of parameters. For those test
 all the sequences were generated randomly, as the target secondary structure.

@@

We built our data set from the 5S rRNA multiple sequence alignment (MSA) available in the Rfam Database 11.0 (Rfam id: \texttt{RF00001}).
Since our software does not currently implement gaps (mainly because scoring indels is a challenging issue that cannot be fully addressed
in this work),  we clustered together the sequences with identical gap locations. From the $54$ MSAs without gap produced, we selected the
biggest MSA  which contains $130$ sequences (out of $712$ in the original Rfam MSA). Then, in order to avoid any bias, we used \texttt{cd-hit}
\cite{Li:2006fk} to remove sequences with more than 80\% of sequence similarity. This operation resulted in a data set of $45$ sequences.
